Simultaneous segmentation and registration of diffusion MR images of the brain driven by active-contours

### 2015-03-12 Publications/2015-NeuroImage/03-introduction.tex
updating paper for second round of reviews

% -*- root: 00-main.tex -*-
\section*{Introduction}\label{sec:introduction}
Nonlinear registration of images is a challenging task, ubiquitous
  in an endless number of image analysis applications.
The process aims to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called
  \emph{reference}, $R$, and \emph{moving}, $M$):%

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}

  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013}
  illustrates the wide variety of existing registration methodologies.
They classify the algorithms by the three principal building blocks that can be identified
  among methods: the matching criteria or similarity function to evaluate the proximity of
  the solution, the deformation model that may implement theoretical properties and constraints
  of the mapping in the selected application, and the optimization method or search strategy.
The possibly most widespread matching criteria are those classified as \emph{iconic
  methods} in \citep{sotiras_deformable_2013}.
These algorithms require the definition of an appropriate function on the voxel-wise information
  content of both images.
A second group, called \emph{geometric methods}, align geometrical features derived from
  corresponding objects defined in both spaces, such as landmarks, shapes, surfaces, etc.
Finally, \emph{hybrid methods} map geometrical features and voxel-wise information.

In this work we propose a hybrid approach to nonlinear registration of anatomically correct
  and precise surfaces to warped and low-resolution \gls*{dmri} of the brain.
Currently, we are witnessing an uprising demand for such algorithms in the quest
  of extracting and characterizing the networks of structural connectivity in the
  human brain \citep{craddock_imaging_2013}.
Our registration method is demonstrated here as a solution to correct for the susceptibility
  distortions that \gls*{dmri} of the head typically present \citep{jezzard_correction_1995}.
The precise segmentation of \gls*{dmri} data is also crucial in such applications, since the
  definition of the nodes in the network is generally derived from a parcellation performed in
  an anatomical \gls*{mri} of the same subject \citep{daducci_connectome_2012}.

\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe  in many different orientations the diffusion process within the living
  brain, these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwith originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations, $\Delta B_0$, is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bones in both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

  \begin{equation}
  \vec{r}' = \vec{r} + \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \cdot \hat{\vec{e}}_{PE}.
  \label{eq:fieldmap}
  \end{equation}

where $\vec{r}$ is the location in the reference domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution was acquiring extra images to estimate $\Delta B_0$ using field mapping
  techniques \citep{jezzard_correction_1995,andersson_modeling_2001}.
A second set of methods aquired extra images reproducing the same \gls*{epi} scheme but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient
  increments in the same axis \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \gls*{dmri}
  data versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
All methods require extra \gls*{mr} aquisitions. Thus, a tipical protocol for the extraction
  of structural connectivity comprehends \gls*{dmri}, \gls*{t1}, and the acquisition demanded
  by the correction method of choice.

Our method is related to the \gls*{t2b} methods as we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to the present approach is found in \citep{greve_accurate_2009}.
They proposed a registration method to align \gls*{t1} and \gls*{dmri} images
  of the brain using a linear transformation model.
Similarly, we exploit the precise structures that can be extracted with
  widely-used software (i.e. \emph{FreeSurfer}~\citep{fischl_freesurfer_2012}),
  and then map these surfaces into warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.
The hypothesis underpinning our research is that, in those situations,
  image registration can be reliably performed by searching for homogeneous
  regions in one image corresponding to a prior knowledge of the shape of objects in
  the moving coordinate system.

>>>>

% -*- root: 00-main.tex -*-
\section{Introduction}\label{sec:introduction}
Image registration is the process to find a spatial mapping $U$ that aligns the information available
  in two different coordinate systems (generally called \emph{reference},
  $R$, and \emph{moving}, $M$):%

  \begin{align}
  U\colon R \subset \mathbb{R}^n &\to M \subset \mathbb{R}^n \notag\\
  \vec{r} &\mapsto \vec{r}' =\vec{r}+u(\vec{r}),
  \label{eq:transform}
  \end{align}
%
  where $\vec{r}$ denotes a position in the reference domain $R$, $\vec{r}'$ is
  its corresponding location in $M$, and $n$ the dimensionality of images.
Finally, $\vec{u} = u(\vec{r})$ is the displacement of every point with respect
  to the reference domain.
A comprehensive survey by \cite{sotiras_deformable_2013} illustrates the wide variety
  of existing methodologies, and classifies them by their three principal building blocks:
  the matching criteria to evaluate the proximity of the solution,
  the deformation model that may implement theoretical properties of the mapping
  (i.e. linearity, elasticity, viscosity, etc.), and the optimization method.
The matching criteria can be defined as a function on the intensities of voxels of the
  two images, the alignment of spatial features derived from the corresponding objects in
  $R$ and $M$, or use both intensity and spatial informations.
In \citep{sotiras_deformable_2013} these methods are referred to as \emph{iconic}, \emph{geometric},
  and \emph{hybrid}, respectively.
When segmentation is introduced to define the matching criteria, we observe a \emph{hybrid}
  registration method that simultaneously defines a partition of the reference space to identify
  relevant objects in it.
An early integration of segmentation and registration by \cite{bertalmio_morphing_2000} proposed
  a sequential deformation of active contours for object tracking in sequential 2D images.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration problem with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration framework using level sets and only
  one \gls*{pde} for first time.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration implementing a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration and segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
\cite{greve_accurate_2009} presented a widely used registration method called \emph{bbregister},
  included in \emph{FreeSurfer} \citep{fischl_freesurfer_2012}.
Their framework performs robust registration of brain surfaces into the intensity information
  of a target \gls*{mri}, using an affine transformation and active contours with edges.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method using level sets and a nonlinear elasticity smoother on the
  displacement vector field, which preserves topology even with very large deformations.
Finally, \cite{gorthi_active_2011} extended the existing methodologies using a multiphase
  level-set function for the registration of several active contours, in the application
  of atlas-based segmentation.
Alternatively to the historical use of active contours based methods, some Bayesian approaches
  have been proposed as well \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.

We propose \emph{regseg}, a registration method based on active contours without edges
  \citep{chan_active_2001} to drive registration and segment the target image.
The hypothesis underpinning our research is that image registration can be reliably performed
  by searching for homogeneous regions in the target image that correspond to precise contours
  from an atlas, or extracted from other image (i.e. a different time step).
We demonstrate the approach in correcting \gls*{dmri} for nonlinear distortions, an application
  that is described in \autoref{sec:distortions}.
In \autoref{sec:methods_map}, we start from a probabilistic framework to finally show the duality
  between the methods described previously.
Following sections introduce
  the specific features of \emph{regseg} (\autoref{sec:numerical_implementation}),
  state the evaluation experiments (\autoref{sec:experiments_evaluation}),
  and describe the synthetic and real data involved in this study (\autoref{sec:datasets}).
In \autoref{sec:results} we describe our findings to support the usefulness of \emph{regseg}
  in the described application, compared to a well-established tool.
The interpretation of the results, and prospects, are addressed in \autoref{sec:discussion}.

### 2015-03-12 Publications/2015-NeuroImage/04-methods.tex
updating paper for second round of reviews

% -*- root: 00-main.tex -*-
\section*{Methods}
\label{sec:methods}
%
\subsection*{Related work}
\label{sec:methods_background}
Registration-segmentation methods are generally built on a deformation model to
  support the evolution of a segmentation algorithm that drives the mapping process.
Thus, solutions usually select first a segmentation strategy to derive the cost function and
  regularization of the problem.
Generally, evolving active contours is the most extended segmentation approach to be
  integrated within the registration process.
These methods either have an implicit description of the relative boundary locations of the
  objects to be delineated using level sets \citep{chen_using_2002,bresson_variational_2006,
  chan_level_2005}, or model the statistical deviations from original shape
  \citep{cremers_kernel_2006,gastaud_combining_2004}.
Some Bayesian approaches have been proposed as well
  \citep{wyatt_map_2003,pohl_bayesian_2006,gass_simultaneous_2014}.
In section \nameref{sec:methods_map}, we start from a probabilistic framework to finally propose
	a cost function defined on the location of explicit surfaces, showing the duality of the
	two existing approaches.

An early ``morphing''-segmentation approach was proposed in \citep{bertalmio_morphing_2000}.
Shortly, \cite{yezzi_variational_2001} presented the first method including a full solution to
  the registration, where the energy functional is defined simultaneously in the target
  and reference images with an affine transformation supporting the coordinates mapping.
\cite{vemuri_joint_2003} proposed an atlas-based registration using a level-sets
  framework and only one \gls*{pde} for first time.
The principal application of these techniques was to perform registration between time steps in
  time series or between different images \citep{paragios_level_2003}.
\cite{unal_coupled_2005} and later \cite{wang_joint_2006},
  extended the ``two \glspl*{pde}'' method of \cite{yezzi_variational_2001}
  to nonlinear registration with a free deformation field.
\cite{droske_mumfordshah_2009} reviewed the latter set of techniques, and proposed two different
  approaches to apply the Mumford-Shah functional \citep{mumford_optimal_1989} in simultaneous
  registration-segmentation, through the propagation of the deformation field from
  the contours to the whole image definition.
Finally, \cite{gorthi_active_2011} presented a comprehensive generalization of the
  existing methodologies based on level sets in the application of atlas-based segmentation.
Recently, \cite{guyader_combined_2011} proposed a simultaneous segmentation and
  registration method close to the one proposed here.
Our method differs on two key features:
  it allows multiple active contours for segmentation, and the regularization strategy.
Whereas \cite{guyader_combined_2011} proposed a nonlinear elasticity smoother, we rely on
  the inherent smoothness of the deformation field, and on the anisotropic regularization
  proposed by \cite{nagel_investigation_1986}.

\subsection*{Cost-function derivation}\label{sec:methods_map}

@@

\subsection*{Numerical Implementation}
\label{sec:numerical_implementation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inner normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	  the vertex $\vec{v}_i$ in each limiting region (see \nameref{app:shape_priors}).
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	  box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer.
	Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	  the corresponding weights $\psi_k(\vec{v}_i)$ \eqref{eq:basis_derivative}.
	}\label{fig:method}
\end{figure}

@@

We can apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to the appendix \nameref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

@@

\subsection*{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
In order to demonstrate the performance of \emph{regseg}, we first conducted a battery of
  accuarcy tests over synthetic data, and then applied it in the susceptibility distortion
  correction of real data.
\autoref{fig:evworkflows} presents the workflow implementing the evaluation instruments.
Besides a visual assessment of the results, we report quantitative evaluations using
  two metrics.
In the case of the phantoms, since we had produced distortions along the three
  available dimensions, we computed the Haussdorf distance, by reusing the
  ``point-to-cell'' method of \cite{commandeur_vtk_2011}.
Conversely, the susceptibility-derived distortions only happen along the \gls*{pe}
  axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $w_i$:

@@

\subsection*{Image Data and preprocessing}
\label{sec:datasets}

@@

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus, $\Gamma_{Tha}$;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures, $\Gamma_{VdGM}$;
  3) cerebral \gls*{wm}, $\Gamma_{WM}$;
  4) brain stem and cerebellar \gls*{wm}, $\Gamma_{Bst}$;
	5) cerebellar \gls*{gm}, $\Gamma_{cbGM}$; and
	6) cortical \gls*{gm} surface, $\Gamma_{pial}$.
The choice of this particular model is further addressed in \nameref{sec:res_model_and_metric}.

>>>>

% -*- root: 00-main.tex -*-
\section{Methods}
\label{sec:methods}
%

\subsection{Susceptibility-derived distortions}
\label{sec:distortions}
\Gls*{dmri} data are usually acquired with \gls*{epi} schemes.
In order to quickly probe the diffusion process within the brain in many different orientations,
  these sequences minimize the bandwidth of the \gls*{pe} direction.
This low bandwidth originates the artifactual displacement of locations presenting small deviations
  from the main magnetic field $B_0$.
The principal source of these deviations ($\Delta B_0$) is the discontinuity of magnetic
  susceptibility at tissue interfaces.
Tissue/air boundaries show large steps of susceptibility and this artifact is
  particularly evident in the surroundings of the orbito-frontal lobe or the temporal
  bone of both hemispheres.
The displacements field $U_s$ along the \gls*{pe} axis can be estimated theoretically
  \citep{jezzard_correction_1995}:

  \begin{equation}
  \vec{r}' = \vec{r} + \frac{\gamma \, T_{acq}\, s_{PE}}{2\pi}\Delta B_0(\vec{r}) \cdot \hat{\vec{e}}_{PE},
  \label{eq:fieldmap}
  \end{equation}
%
where $\vec{r}$ is the true location in the undistorted domain as in \eqref{eq:transform},
  $\gamma$ is the gyromagnetic ratio, $T_{acq}$ is the readout time,
  $s_{PE}$ is the pixel spacing along \gls*{pe}, and $\hat{\vec{e}}_{PE}$ the unitary
  vector along \gls*{pe}.
Several methodologies have been proposed to estimate $U_s$ and fix the artifact.
The first solution consisted on estimating $\Delta B_0$ with field mapping
  techniques \citep{andersson_modeling_2001}.
A second family of methods require an extra \gls*{epi} acquisition, but switching
  the \gls*{pe} axis \citep{chiou_simple_2000} or reversing the direction of gradient \gls*{pe}
  increments \citep{cordes_geometric_2000,holland_efficient_2010}.
\cite{kybic_unwarping_2000} proposed a correction using nonlinear registration of the \emph{b0}
  image\footnote{The term \emph{b0} refers to one or more images within the
  \gls*{dmri} dataset that are acquired with a null or very low gradient intensity or
   \emph{b}-value used as signal reference.} of the \gls*{dmri} versus a \gls*{t2} image.
Further developments \citep{irfanoglu_susceptibility_2011} extended this method to the
  three-dimensional problem.
Therefore, a typical protocol for the extraction of structural connectivity
  comprehends \gls*{dmri}, \gls*{t1}, and the acquisition determined by the correction
  method of choice \citep{daducci_connectome_2012}.

\subsection{Our method and contributions}
\label{sec:our_method}
\emph{Regseg} is related to \emph{bbregister}, as both methods exploit the precise
  surfaces that can be extracted with \emph{FreeSurfer}~\citep{fischl_freesurfer_2012},
  to then map this prior information into a target image.
However, \emph{bbregister} features an affine transform that is unable to compensate the
  distortions.
To avoid this problem, the active contours are dismissed in regions typically affected by
  warpings.
Additionally, \emph{bbregister} is robust in finding clear tissue boundaries.
In the case of \gls*{dmri} data, oftentimes the registration process is restricted to finding
  the pial surface, as it is the only clear boundary.

to the methods based on nonlinear registration of the
  to the \gls*{t2} image, since we also render the correction as a
  registration problem.
Conversely, the method does not require any of the previously mentioned extra acquisitions.
A fundamental antecedent to \emph{regseg} is found in .
They proposed a registration method to align surfaces extracted from the \gls*{t1}
  to \gls*{dmri} images using a linear transformation model using active contours
  with edges.
Similarly, we  warped data, in which the scale of the
  imaged structures is in the range of the image resolution or above.
Our approach differs from \citep{greve_accurate_2009} in two fundamental choices.
First, we use a nonlinear deformation model that enables the correction of the
  susceptibility-derived distortions \citep{jezzard_correction_1995}.
\cite{greve_accurate_2009} circumvented the nonlinearity by identifying typically
  distorted regions and removing them from the metric computation.
Second, our matching criteria does not necessarily need for clear intensity
  steps at the tissue interfaces where surfaces must be fitted, thanks to the use of
  active contours without edges \citep{chan_active_2001}.

This choice is similar to \citep{guyader_combined_2011}, but \emph{regseg} differs in two
  features.
First, the regularization relies on the inherent smoothness of the deformation field,
  and on the anisotropic approach proposed by \cite{nagel_investigation_1986}.
Second, it accepts several contours to be mapped as in \citep{gorthi_active_2011}.
However, \emph{regseg} makes use of the explicit representation of surfaces, and
  does not implement multiphase level sets.
Our registration method is demonstrated here as a solution to correct for the susceptibility
  distortions that \gls*{dmri} of the head typically present \citep{jezzard_correction_1995}.
The distortion correction and the precise segmentation of \gls*{dmri} data is crucial in
	applications such as connectome extraction \citep{craddock_imaging_2013}, since the
  definition of the nodes in the network is generally derived from a parcellation performed in
  an anatomical \gls*{mri} of the same subject \citep{daducci_connectome_2012}.

\subsection{Cost-function derivation}\label{sec:methods_map}

@@

\subsection{Numerical Implementation}
\label{sec:numerical_implementation}

\begin{figure}
	\includegraphics[width=\linewidth]{figures/figure01}
	\caption{The active contours are defined as the interfacing surfaces of the competing
	  \glspl{roi} $\Omega_i$.
	They are represented in green and dark blue colors in this close-up.
	They iteratively evolve following their inner normals $\hat{n}_i$ at each vertex
	  $\vec{v}_i$ of the mesh.
	The gradient speeds $\bar{s}_i$ are computed as the disparity between data energies of
	  the vertex $\vec{v}_i$ in each limiting region (see \autoref{app:shape_priors}).
	In this figure, the gradient speed corresponding to $\vec{v}_1$ is written in the lower
	  box, $\Omega_{wm}$ being the inner limiting region and $\Omega_{gm}$ the outer.
	Finally, every $\bar{s}_i$ is associated with the control points $\vec{u}_k$ through
	  the corresponding weights $\psi_k(\vec{v}_i)$ \eqref{eq:basis_derivative}.
	}\label{fig:method}
\end{figure}

@@

We can apply a discretized interpretation of \eqref{eq:shape_gradients} to compute
  the data term in \eqref{eq:gradient_descent} using shape-gradients
  \citep{herbulot_segmentation_2006}, and ultimately avoid level sets.
We refer the reader to the appendix \autoref{app:shape_priors} in order to fill in the gap
  between \eqref{eq:gradient_descent} and the following expression:

@@

\subsection{Experiments and evaluation}
\label{sec:experiments_evaluation}
%
In order to demonstrate the performance of \emph{regseg}, we first conducted a battery of
  accuarcy tests over synthetic data, and then applied it in the susceptibility distortion
  correction of real data.
\autoref{fig:evworkflows} presents the workflow implementing the evaluation instruments.
Besides a visual assessment of the results, we report quantitative evaluations using
  two metrics.
In the case of the phantoms, since we had produced distortions along the three
  available dimensions, we computed the Haussdorf distance, by reusing the
  ``point-to-cell'' method of \cite{commandeur_vtk_2011}.
Conversely, the susceptibility-derived distortions only happen along the \gls*{pe}
  axis of the image.
Therefore, a \gls*{swindex} can be computed as the one-to-one distance between corresponding
  vertices of surfaces, weighted by their respective Voronoi area $w_i$:

@@

\subsection{Image Data and preprocessing}
\label{sec:datasets}

@@

Selecting the appropriate labels in the \emph{aparc} segmentation, we applied
  the marching-cubes algorithm again to extract the surface of the following
  six homogeneous regions $\Omega_l$:
  1) the thalamus, $\Gamma_{Tha}$;
  2) \gls*{csf} of the ventricular system and deep \gls*{gm} structures, $\Gamma_{VdGM}$;
  3) cerebral \gls*{wm}, $\Gamma_{WM}$;
  4) brain stem and cerebellar \gls*{wm}, $\Gamma_{Bst}$;
	5) cerebellar \gls*{gm}, $\Gamma_{cbGM}$; and
	6) cortical \gls*{gm} surface, $\Gamma_{pial}$.
The choice of this particular model is further addressed in \autoref{sec:res_model_and_metric}.
